A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2023-12-31 paper/paper.tex
Remove stlc review section

The key technique that underlies our consistency proof is proof by
logical relation. In Section~\ref{sec:logrelstlc}, we review the
technique in the context of the simply typed lambda calculus. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary.

@@

\section{Review of Proof by Logical Relation}
\label{sec:logrelstlc}
In this section, we give a quick review of the consistency proof of the
simply typed lambda calculus augmented with the empty type through
proof by logical relation.

We give the full specification of the simply typed calculus in
Figure~\ref{fig:stlcsyntax},\ref{fig:stlctyping},\ref{fig:stlcdynamics}
without any further explanation since it is standard.

The goal here is to show that the empty type cannot be
inhabited. A naive attempt at proof by structural induction

@@

\begin{lemma}[ST-Pi]
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. Applying
% Our goal is to show that there exists some
%   $[[S]]$ such that $[[InterpR i Pi A B S ]]$.
%   From the premise $[[G |= A : Set i]]$, we know that there exists
%   some $[[S0]]$ such that
\end{proof}

>>>>

The key technique that underlies our consistency proof is proof by
logical relation. In
Section~\ref{sec:spec}, we present the dependent type theory of
interest. In Section~\ref{sec:logreldep}, we give the definition
of the logical relation for the dependent type theory. Rather than
presenting the logical relation as an inductive-recursive definition,
we use the more elementary concept of a partial function to capture
the interpretation of types. The alternative representation requires us
to show that the set of equations indeed defines a partial function;
that is, for each input, there should always be a unique
output.
From the interpretation function, we can define the semantic
typing judgment for the set of lambda terms.
In Section~\ref{sec:logrelproof}, we prove the fundamental lemma,
which states that syntactic typing implies semantic typing. Once the
fundamental lemma is established, logical consistency follows as a
trivial corollary.

@@

@@

\begin{lemma}[ST-Pi]
  If $[[G |= A : Set i]]$ and $[[G ++ A |= B : Set i]]$, then $[[G |= Pi
  A B : Set i]]$.
\end{lemma}
\begin{proof}
  Applying Lemma~\ref{lemma:setinv} to the
  conclusion, it now suffices to show that given $[[rho |= G]]$, there
  exists some $[[S]]$ such that $[[InterpR i Pi A B S]]$.
  From Lemma~\ref{lemma:setinv} and $[[G |= A : Set i]]$, we know that
  there exists some set $[[S0]]$ such that $[[InterpR i A S0]]$.
  Let $[[F]]$ be the following relation:
  \[ (a, S) \in [[F]] \iff [[InterpR i B {rho .: a} S]] \]
  By Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a partial function with
  the signature $[[STm -> PowerSet STm]]$.

We
  instantiate $[[S0]]$ with the set $\{ b \mid \forall [[ a in S0]],
  \forall [[S1]], [[InterpR i B {a} S1]] \implies [[b a in S1]] \}$.
\end{proof}
